Microbiome and hypertension

### 2018-12-19 rrbiome.Rmd
Beta diversity

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, results='asis', cache=TRUE, warning=FALSE)
knitr::opts_chunk$set(cache.path = 'cache/', output.dir="cache/",
                      file.path = 'cache/', fig.path = 'cache/')
```

Importing libraries

```{r libraries}

library(dplyr)
library(tibble)
library(phyloseq)
library(microbiome)
library(finriskmetagcommon)
library(knitr)
library(tidyr)
library(vegan)

```

```{r functions, include = FALSE}

print.dset <- function(dset) {
    dset %>%
        gather(Covariate, Value) %>%
        group_by(Covariate) %>%
        summarise_all(., .funs = funs(mean(as.numeric(.), na.rm = TRUE),
                                         median(as.numeric(.), na.rm = TRUE),
                                         sd(as.numeric(.), na.rm = TRUE),
                                         min(as.numeric(.), na.rm = TRUE),
                                         max(as.numeric(.), na.rm = TRUE),
                                         NAs = sum(is.na(.)),
                                         n = n(),
                                         n_distinct)) %>%
            mutate_if(is.numeric, round, 2) %>%
            kable
}

```

Loading phyloseq object

```{r data}

pseq <- readRDS("data/phfinrisk_genus_all_drop50k_2018-11-16.RDs")

```

Loading descriptions for clinical data

```{r variables}

if (!exists('phfinrisk_metadatadesc')) {
  # this is a hand-compiled data.frame that contains metadata information for variables utilized in our analyses
  data("phfinrisk_metadatadesc", package="finriskmetagcommon")
}
names.dset <- filter(phfinrisk_metadatadesc, Ignored.Covariate.in.Cross.Sectional.Analysis.Aaro20181115==0)

```

Filtering clinical data

```{r filtering}

dset <- meta(pseq) %>%
    tibble::rownames_to_column(var = "rowname") %>%
    dplyr::select(rowname, DNA_OK, MEN, BL_AGE, SYSTM, DIASM, BP_TREAT, BMI, CURR_SMOKE, DIAB, BL_USE_RX_C09,
        Q57X, BL_USE_RX_C03, BL_USE_RX_C07, BL_USE_RX_C08) %>%
    dplyr::mutate(PULSEPRESSURE = SYSTM - DIASM,
                  HYPERTENSION = factor(ifelse(SYSTM >= 140 | DIASM >= 90 | BP_TREAT == 1, 1, 0)),
                  SEX = factor(ifelse(MEN == "Female", 1, 0)),
                  MAP = 2./3.*DIASM + 1./3.*SYSTM) %>%
    dplyr::select(-MEN) %>%
    stats::na.omit() %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames(var = "rowname")

print.dset(dset)

```

Continuous variables

```{r distribution plots, fig.width=10, fig.height=4}
ggplot((dset %>%
        select(BL_AGE, BMI, SYSTM, DIASM, PULSEPRESSURE, MAP) %>%
        gather(Covariate, Value)), aes_string(x = "Value")) +
    facet_wrap(~Covariate, ncol=4, scales = "free") +
    geom_density(aes(y=..scaled..)) + 
    ggplot2::theme_minimal() 
```

Categorical variables

```{r histogramsy, fig.width=10, fig.height=4}
ggplot((dset %>%
        select(-BL_AGE, -BMI, -SYSTM, -DIASM, -PULSEPRESSURE, -MAP) %>%
        gather(Covariate, Value)), aes_string(x = "Value")) +
    facet_wrap(~Covariate, ncol=8, scales = "free") +
    geom_bar() + 
    ggplot2::theme_minimal() 
```

```{r Updating meta data}

pseq_filtered <- pseq
sample_data(pseq_filtered) <- sample_data(dset)

```

# Alpha diversity

Dset with alpha diversity

```{r alpha diversity}

alphadiversity <- microbiome::global(pseq_filtered, index="shannon")
dset.alpha <- merge(dset, alphadiversity, by=0, all=TRUE)

```

```{r alpha diversity scatterplot, fig.width=8, fig.height=4}

ggplot(dset.alpha, aes(x=diversities_shannon, y=MAP)) +
    geom_point(size=1, shape=1) +
    coord_fixed() +
    ggplot2::theme_minimal() 
```

Linear models

```{r lm model}
lm(MAP ~ diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)
lm(MAP ~ BL_AGE + SEX + diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)
lm(MAP ~ BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + BP_TREAT + diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)

```

# Beta diversity

Matriisin teko _bc_dissimilarity_matrix.R ja käyttö comp_adonis_covariates.R

```{r vegdist}

phf_profile.rel <- microbiome::transform(pseq_filtered, 'compositional')
otu <- microbiome::abundances(phf_profile.rel)

if (!file.exists("rds/bray.dist.m.rds")) {
    bray.dist.m <- vegdist(t(otu), method="bray")
    saveRDS(bray.dist.m, file = "rds/bray.dist.m.rds")
} else {
    bray.dist.m <- readRDS("rds/bray.dist.m.rds")
}

```

```{r adonis}

attr(bray.dist.m, "method")
dist.matrix <- as.matrix(bray.dist.m)
attr(dist.matrix, "method") <- "bray"

if (!file.exists("rds/phf.adonis.rds")) {
phf.adonis <- adonis(dist.matrix ~ BL_AGE + SEX + MAP,
                     data = meta(pseq_filtered),
                     permutations = 99)
    saveRDS(phf.adonis, file = "rds/phf.adonis.rds")
} else {
    phf.adonis <- readRDS("rds/phf.adonis.rds")
}

phf.adonis$aov.tab %>% as.data.frame %>% kable

```

```{r beta heatmap, eval = FALSE}

ggplot(data = dist.matrix, aes(x = metrics, y = teams)) +
      geom_tile(aes(fill = performance))

```
